---
title: "LQ-PCA"
author: "Ryan Thomas"
date: "June 23, 2016"
output: html_document
---
# About the data
```{r setup }
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(rgdal)
library(sp)
library(dplyr)
library(data.table)
```

```{r getData}
counties <- readOGR("https://datahub.io/dataset/e82a4e76-ccf0-4d3a-b2c8-eda2b52ab32a/resource/f2693c94-fd69-4baf-887b-f302a5188ead/download/counties.geojson",
                    "OGRGeoJSON")
counties@data$FIPS <- as.character(counties@data$FIPS)
for (i in 1:nrow(counties@data) ) {
  if ( nchar( counties@data$FIPS[i] ) == 4) {
    counties@data$FIPS[i] <- paste("0",as.character(counties@data$FIPS[i]), sep = "")
  } else {
     counties@data$FIPS[i] <- counties@data$FIPS[i]
  }
}
counties@data$FIPS <- as.factor(counties@data$FIPS)
str(counties@data$FIPS)
counties@data <- counties@data[,1:10]

sectors <- data.frame(cbind(
  c("11",	"21",	"22",	"23",	"31_33",	"42",	"48_49",	
    "51",	"52",	"53",	"54",	"55",	"56",	"61", "62","71","72","81","92"),
  c("AgriForest",	"Mining",	"Utilities",	"Construction",	
    "Manufacturing",	"WholeTrade", "RetailTrade",	"Transportation",	"Information",	
    "Finance",	"RealEstate",	"ProfScienceTech",	
    "Management",	"AdminWasteManagement","Education","HealthCare","AccomodationFood","Other","PubAdmin")))
names(sectors) <- c("sector", "description")

qcewGetIndustryData <- function (year, qtr, industry) {
  url <- "http://www.bls.gov/cew/data/api/YEAR/QTR/industry/INDUSTRY.csv"
  url <- sub("YEAR", year, url, ignore.case=FALSE)
  url <- sub("QTR", qtr, url, ignore.case=FALSE)
  url <- sub("INDUSTRY", industry, url, ignore.case=FALSE)
  read.csv(url, header = TRUE, sep = ",", quote="\"", dec=".", na.strings=" ", skip=0)
}
### Use this loop to go through the sectors loaded in the data frame 
for (i in 1:length(sectors[,1])) {
  if (i == 1) {
    print(paste("getting data for ", sectors[i,2]))
    dt1 <- qcewGetIndustryData("2015", "3", sectors[1,1])
    dt1 <- aggregate(dt1$month2_emplvl, by=list(Category=dt1$area_fips), FUN=sum)
    names(dt1) <- c("area_fips", as.character(sectors[1,2]))
    
  } else if (i == 2) {
    print(paste("getting data for ", sectors[i,2]))
    dt2 <- qcewGetIndustryData("2015", "3", sectors[i,1])    
    dt2 <- aggregate(dt2$month2_emplvl, by=list(Category=dt2$area_fips), FUN=sum)
    names(dt2) <- c("area_fips", as.character(sectors[i,2]))
    data <- merge(dt1, dt2, all=TRUE)
    
  } else {
    print(paste("getting data for ", sectors[i,2]))
    dt3 <- qcewGetIndustryData("2015", "3", sectors[i,1]) 
    dt3 <- aggregate(dt3$month2_emplvl, by=list(Category=dt3$area_fips), FUN=sum)
    names(dt3) <- c("area_fips", as.character(sectors[i,2]))
    data <- merge(data, dt3, all=TRUE)
    head(data)
      }
  
}
```

#### Assign row names and column names for reference through the transformations

```{r rename}
rownames(data) <- data$area_fips
colnames(data) <- c("FIPS", "AgriForest",	"Mining",	"Utilities",	"Construction",	
    "Manufacturing",	"WholeTrade", "RetailTrade",	"Transportation",	"Information",
    "Finance",	"RealEstate",	"ProfScienceTech", "Management",	"AdminWasteManagement",
    "Education","HealthCare", "AccomodationFood","Other","PubAdmin")
glimpse(data)
str(data)
```

```{r calculateLQ}
#this can only be computed for complete cases
data <- data[complete.cases(data),]

## cut out the US000 row, which is total US per sector
data <- data[data$FIPS != "US000",]

### Convert the data frame to a matrix and take a look
mat <- data.matrix( data[ ,2:length(data[1,]) ] )

# Calculate the LQ
share_tech_city <- mat / rowSums (mat)
share_tech_city

share_tech_total <- colSums (mat) / sum (mat)
share_tech_total

lq <- t(share_tech_city) / share_tech_total
lq[is.na(lq)] <- 0 
lq[which(lq==Inf)] <- 0 
lq <- t(lq)


# here we go, this is the location quotient for each spatial unit - economic unit pair
colnames(lq) <- list("AgriForest",	"Mining",	"Utilities",	"Construction",	
    "Manufacturing",	"WholeTrade", "RetailTrade",	"Transportation",	"Information",
    "Finance",	"RealEstate",	"ProfScienceTech", "Management",	"AdminWasteManagement",
    "Education","HealthCare", "AccomodationFood","Other","PubAdmin")
head(lq)
```


Data were downloaded from the Bureau of Labor Statistics, and they contain the employment levels for each county at the 2-digit NAICS level. This same analysis could be replicated at various other levels of analysis by adapting the R script that lists the NAICS codes. More information on the BLS API and data availability can be found at the [BLS website](enter here!!! )


## Factor Analysis with Verimax Rotation of LQ data
Factor analysis provides insight to the underlying relationship between multiple variables in a data set of numeric data. For this post, we analyzed the location quotients of each county in the US to see what patterns would emerge in the co-occurence of sectors. In the output below, each factor represents a composite variable based on the loadings shown in the column. Using these loadings, the factors were applied to each county using matrix multiplication to determine a score for each factor shown. The resulting score was then attributed to that county as the factor score for that variable. This method essentially simplifies the economy into 5 composite sectors. Omitted columns represent insignificant loadings, and these were also removed from the matrix multiplication. 
```{r factanal}
##########################################################
##  Perform Factor Analysis using Maximum Likelihood
##  with Varimax Rotation
##########################################################

# These are already complete cases
#lq <- lq[complete.cases(lq),]

fact1=factanal(lq[,c(1:17,19)],factors=5,rotation="varimax")
fact1

#get loading plot for factors 1 by 2-5
par(mfrow=c(3,3),mar = c(4,4,1,1))
plot(fact1$loadings, pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings, labels=colnames(lq),xpd=TRUE)

#get loading plot for factors 1 through 5
plot(fact1$loadings[,c(1,3)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(1,3)], labels=colnames(lq),xpd=TRUE)

plot(fact1$loadings[,c(1,4)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(1,4)], labels=colnames(lq),xpd=TRUE)

plot(fact1$loadings[,c(1,5)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(1,5)], labels=colnames(lq),xpd=TRUE)
# 4

plot(fact1$loadings[,c(2,3)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(2,3)], labels=colnames(lq),xpd=TRUE)

plot(fact1$loadings[,c(2,4)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(2,4)], labels=colnames(lq),xpd=TRUE)

plot(fact1$loadings[,c(2,5)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(2,5)], labels=colnames(lq),xpd=TRUE)

plot(fact1$loadings[,c(3,4)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(3,4)], labels=colnames(lq),xpd=TRUE)

plot(fact1$loadings[,c(3,5)], pch=18, col='red')
abline(h=0)
abline(v=0)
text(fact1$loadings[,c(3,5)], labels=colnames(lq),xpd=TRUE)
#8

#get reproduced correlation matrix
repro1=fact1$loadings%*%t(fact1$loadings)
#residual correlation matrix -- should be close to zero
resid1=fact1$cor-repro1
round(resid1,2)

#get root-mean squared residuals
len=length(resid1[upper.tri(resid1)])
RMSR1=sqrt(sum(resid1[upper.tri(resid1)]^2)/len)
RMSR1

#get proportion of residuals greater than 0.05 in absolute value
sum(rep(1,len)[abs(resid1[upper.tri(resid1)])>0.05])/len
# smaller is better -- 0.2549 is ok but not great
```

## Factor Characterizations
To make these factors understandable, we will go through each factor, look at the loadings and name them based on what they represent.
####Factor 1: 
The first factor has relatively high scores in Produducing and Natural Resources, and relatively low scores in Services. So counties with high scores in this factor would be 

```{r createFactCols }
fal <- matrix(fact1$loadings[c(1:18),],nrow=18, ncol=5, dimnames = list(c("AgriForest",	"Mining",	"Utilities",	"Construction",	
    "Manufacturing",	"WholeTrade", "RetailTrade",	"Transportation",	"Information",	"Finance",	"RealEstate",	"ProfScienceTech",	
    "Management",	"AdminWasteManagement","Education","HealthCare","AccomodationFood","PubAdmin"), 
    c("Factor1", "Factor2",   "Factor3",   "Factor4",    "Factor5")))

for (i in 1:nrow(lq)) {
  
   lq$fact1[i] <- sum(data.frame(
     mapply(`*`,lq[ i, c(5,6,8:13,15,17,18)], fal[c(5,6,8:13,15,17,18),1],SIMPLIFY=FALSE))
     )
   ############
   lq$fact2[i] <- sum(data.frame(
     mapply(`*`,lq[ i, c(4:7,9:12,14,15)], fal[c(1:4,6:9,11,12),2],SIMPLIFY=FALSE))
     )
   lq$fact3[i] <- sum(data.frame(
     mapply(`*`,lq[ i, c(4:9,14,16)], fal[c(1:6,11,13),3],SIMPLIFY=FALSE))
     )
   lq$fact4[i] <- sum(data.frame(
     mapply(`*`,lq[ i, c(4,6,7,12,13)], fal[c(1,3,4,9,10),4],SIMPLIFY=FALSE))
     )
   lq$fact5[i] <- sum(data.frame(
     mapply(`*`,lq[ i, c(4,5,7,14,15)], fal[c(1,2,4,11,12),5],SIMPLIFY=FALSE))
     )
}

lqdf <- as.data.frame(lq)
str(lq)


lqdf$FIPS <- rownames(lqdf)
str(lqdf)
```


```{r mergeCounties}
setkey(counties@data,FIPS)
setkey(lqdf ,FIPS)
counties@data <- merge(x = counties@data, y = lq, all.x = TRUE)
str(counties@data)
```

[leaflet link](https://rstudio.github.io/leaflet/showhide.html)
```{r leaflet, warning=FALSE }
library(rgdal)
library(leaflet)
library(shiny)

pal <- colorQuantile("YlGn", NULL, n = 5)

county_popup <- paste0("<strong>County: </strong>", 
                      counties$NAME, 
                      "<br><strong>Factor 1 Score: </strong>", 
                      round(counties$fact1),2)

leaflet() %>%
  setView(lng = -90, lat = 37, zoom = 3) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = counties, 
              group = "factors",
              fillColor = ~pal(fact1),
              stroke = FALSE,
              fillOpacity = 0.6, 
              color = "#BDBDC3", 
              weight = 1, 
              popup = county_popup)

```

