---
title: "Quito air quality - modeled on Beijing Air Quality"
author: "Moroney & Thomas"
date: "Sep 21, 2016"
output: html_document
---

``````{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      warning=FALSE, message=FALSE)
```

```{r}
library(lubridate)
library(plotrix)
library(dplyr)
library(animation)
aq <- read.table("~/code/urban-epi/air_files/air_quality.tsv", header=TRUE, sep="\t", as.is=T, na.strings = "NA")

# abbreviated month name in spanish? need to set sys locale to understand
Sys.setlocale(locale="es_ES.utf8")

#note 432 failed to parse
aq$date <- as.POSIXct(parse_date_time(aq$yyyy.mm.dd.hh,"%Y-%b-%d-%H"))
aq$yrmonth <- as.factor(paste(year(aq$date),month(aq$date),sep="-"))
aq$year <- as.factor(year(aq$date))
aq$month <- as.factor(month(aq$date))
aq$hour <- as.factor(hour(aq$date))
str(aq)

# drop last three stations from analysis because of all NAs

a <- aq

## omit the ND to find lowest value to set ND to half

# make chemical type a factor
a$Type <- as.factor(a$Type)

# make numeric from character, creates nas
a[, 3:14] <- sapply(a[, 3:14], as.numeric) 

# make summary of data by each
lapply(split(a[, 3:14], a$Type), summary, na.rm=T)

# summary shows that minimum value for all pollutants in each station is 0 (except CO at Los.Chillos 0.04). Therefore, I suggest just dropping all NAs, NDs for analysis

# row means for each
a$mean <- apply(a[, 3:14], 1, mean, na.rm =T)

#For each sensor, let’s figure out how many readings they’ve made and when they are active.
#- You can present this in a simple excel or text doc just by looking at the data, if that's easier than trying to get R to write this #out.

# histogram of dates for each sensor, histogram facetting by sensor for each month



# Then we should count daily exceedances for each sensor and the average. 
# make category for each group of exceedances


```


```{r}
aq_by_hour <- aggregate(cbind(PM2.5.level, Concentration) ~ hour, 
                         data = aq, mean)
aq_by_month <- aggregate(cbind(PM2.5.level, Concentration) ~ yrmonth, 
                         data = aq, mean)

aq_2_2016 <- aggregate(cbind(PM2.5.level, Concentration) ~ hour, 
                       data = aq[aq$date > "2016-02-01" & aq$date < "2016-03-01" ,], mean)

year <- levels(aq$year)
month <- levels(aq$month)

aq_monthly_by_hour <- aggregate(cbind(PM2.5.level, Concentration) ~ year+month+hour, 
                       data = aq, mean)
```

### PM2.5 level in February 2016 was the lowest since January 2011.
PM2.5 levels and concentrations shift throughout the course of the day. This can be seen in the figures below, where the highest mean PM2.5 levels and concentrations occur between 11PM and Midnight, following the evening rush hour. The lowest levels and concentrations occur between 2PM an 4PM. The mean PM2.5 level in January 2016 was the lowest since February 2011. In the middle graph below, the hourly average for Feburary 2016 can be seen in green, clearly below the mean over the period 2011-2016. 
```{r}
par(mfrow = c(1,3), mar = c(4,3,1,1))
polar.plot(aq_by_hour$PM2.5.level,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250), # defaults to min and max
           # main = paste("Mean PM 2.5 and Concentration", "at each Hour in Beijing", sep = "\n"))
           main = "Mean PM2.5 and Concentration at each Hour")
polar.plot(aq_by_hour$Concentration, 
           start = 90, 
           clockwise = T, 
           rp.type = "polygon",
           line.col = "red", 
           show.grid.labels = 1,
           labels = aq_by_hour$hour,
           label.pos = seq(0,360, by = 15),
           radial.lim = c(0, max(aq_by_hour$Concentration)),
           add = TRUE)
legend("bottomright", lty = 1, col = c("blue", "red"), 
       legend = c("PM2.5", "Concentration"),
       box.lwd = 0)
###########################################################################
polar.plot(aq_by_hour$PM2.5.level,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250),
           # main = paste("Mean PM 2.5 and Concentration", "at each Hour in Beijing", sep = "\n"))
           main = "Monthly Mean PM2.5 at each Hour")
for (i in year){
  for (j in month){
    polar.plot(aq_monthly_by_hour$PM2.5.level[aq_monthly_by_hour$month == j & 
                                                aq_monthly_by_hour$year == i], 
               start = 90, 
               clockwise = T, 
               rp.type = "polygon",
               line.col = "lightgrey", 
               show.grid.labels = 1,
               labels = aq_monthly_by_hour$hour,
               label.pos = seq(0,360, by = 15),
               radial.lim = c(0, 250),
               lwd = 0.2,
               add = TRUE)
    }
  }

polar.plot(aq_by_hour$PM2.5.level,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250), # defaults to min and max
           add = TRUE)
polar.plot(aq_monthly_by_hour$PM2.5.level[aq_monthly_by_hour$month == "2" & aq_monthly_by_hour$year == "2016"], 
           start = 90, 
           clockwise = T, 
           rp.type = "polygon",
           line.col = "green", 
           show.grid.labels = 1,
           labels = aq_monthly_by_hour$hour,
           label.pos = seq(0,360, by = 15),
           radial.lim = c(0, max(aq_by_hour$PM2.5.level)),
           add = TRUE)
legend("bottomright", lty = 1, col = c("lightgrey", "blue", "green"), 
       legend = c("PM2.5 level each month", "Mean", "Feb 2016"),
       box.lwd = 0)
###########################################################################
polar.plot(aq_by_hour$Concentration,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250),
           # main = paste("Mean PM 2.5 and Concentration", "at each Hour in Beijing", sep = "\n"))
           main = "Monthly Mean Concentration at each Hour")
for (i in year){
  for (j in month){
    polar.plot(aq_monthly_by_hour$Concentration[aq_monthly_by_hour$month == j & 
                                                aq_monthly_by_hour$year == i], 
               start = 90, 
               clockwise = T, 
               rp.type = "polygon",
               line.col = "lightgrey", 
               show.grid.labels = 1,
               labels = aq_monthly_by_hour$hour,
               label.pos = seq(0,360, by = 15),
               radial.lim = c(0, 250),
               lwd = 0.2,
               add = TRUE)
  }
}

polar.plot(aq_by_hour$Concentration,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "red", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250), # defaults to min and max
           add = TRUE)
polar.plot(aq_monthly_by_hour$Concentration[aq_monthly_by_hour$month == "2" & aq_monthly_by_hour$year == "2016"], 
           start = 90, 
           clockwise = T, 
           rp.type = "polygon",
           line.col = "black", 
           show.grid.labels = 1,
           labels = aq_monthly_by_hour$hour,
           label.pos = seq(0,360, by = 15),
           radial.lim = c(0, max(aq_by_hour$PM2.5.level)),
           add = TRUE)
legend("bottomright", lty = 1, col = c("lightgrey", "red", "black"), 
       legend = c("PM2.5 con. each month", "Mean", "Feb 2016"),
       box.lwd = 0)
```


### Mean PM2.5 Level by Month
```{r}
aq_by_month <- aq_by_month %>%
    mutate(yrmonth =  factor(yrmonth, levels = yrmonth_order)) %>%
   arrange(yrmonth)
plot(aq_by_month$PM2.5.level,
     type = "b",
     axes = FALSE,
     xlab = "Year-Month",
     ylab = "PM2.5",
     main="Mean PM 2.5 Level by Month",
     ylim = c(0,250),
     pch = 16,
     col = "grey")
  axis(1, at= seq(1,73,by=2), 
       tck = FALSE,
       labels = FALSE)
  text(seq(1,73,by=2), par("usr")[3]-0.5, 
       srt = 60, adj= 1, xpd = TRUE,
       labels = paste(levels(aq_by_month$yrmonth)[seq(1,73,by=2)]), 
       cex=0.8)
  axis(2, at= seq(0,(max(aq_by_month$PM2.5.level)+50),by=50),
       tck = 0, las = 1,
       labels = seq(0,(max(aq_by_month$PM2.5.level)+50),by=50))
  par(xpd = FALSE)
  grid()
   points(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2016-2"],
        y = min(aq_by_month$PM2.5.level),
        col = "red",
        pch = 16)
   points(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2011-1"],
         y = aq_by_month$PM2.5.level[aq_by_month$yrmonth=="2011-1"],
         col = "blue",
         pch = 16)
   text(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2015-10"],
       y = min(aq_by_month$PM2.5.level)-20,
       paste("PM2.5 = ", round(min(aq_by_month$PM2.5.level),2)," in Feb 2016", sep = "\n"),
       adj = c(0,1),
       offset = c(0,-2))
   text(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2011-2"],
       y = aq_by_month$PM2.5.level[aq_by_month$yrmonth=="2011-1"]-20,
       paste("PM2.5 = ", round(aq_by_month$PM2.5.level[aq_by_month$yrmonth=="2011-1"],2), 
             "in Jan 2011", sep = "\n"),
       adj = c(0,1))
```


