---
title: "Quito air quality - modeled on Beijing Air Quality"
author: "Moroney & Thomas"
date: "Sep 21, 2016"
output: html_document
---

``````{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8,
                      warning=FALSE, message=FALSE)
```

```{r}
library(zoo)
library(lubridate)
#library(plotrix)
library(dplyr)
#library(animation)
library(tidyr)
library(ggplot2)

## READ IN DATA
aq <- read.table("~/code/urban-epi/air_files/air_quality.tsv", header=TRUE, sep="\t", as.is=T, na.strings = "NA")

## COLUMN VARIABLES TO CORRECT CLASS
# date to posix
# abbreviated month name in spanish? need to set sys locale to understand
Sys.setlocale(locale="es_ES.utf8")
aq$date <- as.POSIXct(parse_date_time(aq$yyyy.mm.dd.hh,"%Y-%b-%d-%H")) #432 failed to parse
# chemical type to factor
aq$Type <- as.factor(aq$Type)
# make conc values numeric from character. 
aq[, 3:14] <- sapply(aq[, 3:14], as.numeric) #creates some NAs from the NDs, #last three stations all NAs?
# make new factors out of date data
#aq$yrmonth <- as.factor(paste(year(aq$date),month(aq$date),sep="-"))
#aq$year <- as.factor(year(aq$date))
#aq$month <- as.factor(month(aq$date))
#aq$hour <- as.factor(hour(aq$date))

# prove 12:14 have all NAs with summary, shows all NAs for those three, but when using dplyr they aren't?
lapply(split(aq[, 3:14], aq$Type), summary, na.rm=T)

## TIDY DATA
# new df - make station columns turn into rows, drop three stations that are all NAs
a <- aq %>%
	gather(station, conc, 3:11)%>%
	rename(chem = Type) %>%
	select(date, station, chem, conc) 

a$station <- as.factor(a$station)

## ANALYSIS
# number of missing obs (31%)
sum(is.na(a$conc))/length(a$conc)
# number of obs for each Type
lapply(split(aq[, 3:12], aq$Type), function(x) sum(!is.na(x)))
# how many obs per station, showing 0 non NAs for El Condado?
apply(aq[, 3:12], 2, function(x) sum(!is.na(x)))

# make summary of data by each, drop NAs
a[!is.na(a$conc),] %>% 
	group_by(station, chem) %>%
	summarize(mean_conc = mean(conc), max_conc = (max(conc)), n = n()) %>%
	arrange(chem, desc(mean_conc))

# row means for each
aq$mean <- apply(aq[, 3:14], 1, mean, na.rm = T)

# when they are active
a[!is.na(a$conc),] %>% 
	group_by(station) %>%
	summarize(start = date[1], end = date[n()])

# histogram of dates for each sensor
ggplot(a, aes(x = date)) + 
	geom_histogram(bins=98) + #8yr*12mo 
	facet_wrap(~station)

# histogram facetting by sensor for each month
a %>% 
	mutate(month = month(date)) %>%
	ggplot(aes(x = month)) + 
	geom_histogram() + 
	facet_wrap(~station)
```

```{r}
# Then we should count daily exceedances for each sensor and the average. 
# make category for each group of exceedances

# CO need 8 hour avg (ug/m3), units different from legis? max value is 11.7 from Carapungo but lowest legislated value is 15000
a[!is.na(a$conc),] %>% 
	filter(chem == "CO") %>%
	mutate(CO.8 = rollmean(x = conc, 8, align = "right", fill = NA)) %>% 
	group_by(station) %>% 
	summarize(Alert_ex = sum(conc > 1.5E4) , Alarm_ex = sum(conc > 3E4), Emergy_ex = sum(conc > 4E4))

# O3 avg in 8 hours
a[!is.na(a$conc),] %>% 
	filter(chem == "O3") %>%
	mutate(O3.8 = rollmean(x = conc, 8, align = "right", fill = NA)) %>%
	group_by(station) %>%
	summarize(Alert_ex = sum(conc > 2E2), Alarm_ex = sum(conc > 4E2), Emergy_ex = sum(conc > 6E2))

# PM 2.5 in 24 hours
a[!is.na(a$conc),] %>% 
	filter(chem == "PM25") %>%
	mutate(PM25.24 = rollmean(x = conc, 24, align = "right", fill = NA)) %>%
	group_by(station) %>%
	summarize(Alert_ex = sum(conc>1.5E2), Alarm_ex = sum(conc>2.5E2))

# SO2 in 24 hours
a[!is.na(a$conc),] %>% 
	filter(chem == "CO") %>%
	mutate(SO2.24 = rollmean(x = conc, 24, align = "right", fill = NA)) %>%
	group_by(station) %>%
	summarize(Alert_ex = sum(conc>2E2), Alarm_ex = sum(conc>1E3), Emergy_ex = sum(conc>1.8E3))

```


```{r}
aq_by_hour <- aggregate(cbind(PM2.5.level, Concentration) ~ hour, 
                         data = aq, mean)
aq_by_month <- aggregate(cbind(PM2.5.level, Concentration) ~ yrmonth, 
                         data = aq, mean)

aq_2_2016 <- aggregate(cbind(PM2.5.level, Concentration) ~ hour, 
                       data = aq[aq$date > "2016-02-01" & aq$date < "2016-03-01" ,], mean)

year <- levels(aq$year)
month <- levels(aq$month)

aq_monthly_by_hour <- aggregate(cbind(PM2.5.level, Concentration) ~ year+month+hour, 
                       data = aq, mean)
```

### PM2.5 level in February 2016 was the lowest since January 2011.
PM2.5 levels and concentrations shift throughout the course of the day. This can be seen in the figures below, where the highest mean PM2.5 levels and concentrations occur between 11PM and Midnight, following the evening rush hour. The lowest levels and concentrations occur between 2PM an 4PM. The mean PM2.5 level in January 2016 was the lowest since February 2011. In the middle graph below, the hourly average for Feburary 2016 can be seen in green, clearly below the mean over the period 2011-2016. 
```{r}
par(mfrow = c(1,3), mar = c(4,3,1,1))
polar.plot(aq_by_hour$PM2.5.level,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250), # defaults to min and max
           # main = paste("Mean PM 2.5 and Concentration", "at each Hour in Beijing", sep = "\n"))
           main = "Mean PM2.5 and Concentration at each Hour")
polar.plot(aq_by_hour$Concentration, 
           start = 90, 
           clockwise = T, 
           rp.type = "polygon",
           line.col = "red", 
           show.grid.labels = 1,
           labels = aq_by_hour$hour,
           label.pos = seq(0,360, by = 15),
           radial.lim = c(0, max(aq_by_hour$Concentration)),
           add = TRUE)
legend("bottomright", lty = 1, col = c("blue", "red"), 
       legend = c("PM2.5", "Concentration"),
       box.lwd = 0)
###########################################################################
polar.plot(aq_by_hour$PM2.5.level,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250),
           # main = paste("Mean PM 2.5 and Concentration", "at each Hour in Beijing", sep = "\n"))
           main = "Monthly Mean PM2.5 at each Hour")
for (i in year){
  for (j in month){
    polar.plot(aq_monthly_by_hour$PM2.5.level[aq_monthly_by_hour$month == j & 
                                                aq_monthly_by_hour$year == i], 
               start = 90, 
               clockwise = T, 
               rp.type = "polygon",
               line.col = "lightgrey", 
               show.grid.labels = 1,
               labels = aq_monthly_by_hour$hour,
               label.pos = seq(0,360, by = 15),
               radial.lim = c(0, 250),
               lwd = 0.2,
               add = TRUE)
    }
  }

polar.plot(aq_by_hour$PM2.5.level,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250), # defaults to min and max
           add = TRUE)
polar.plot(aq_monthly_by_hour$PM2.5.level[aq_monthly_by_hour$month == "2" & aq_monthly_by_hour$year == "2016"], 
           start = 90, 
           clockwise = T, 
           rp.type = "polygon",
           line.col = "green", 
           show.grid.labels = 1,
           labels = aq_monthly_by_hour$hour,
           label.pos = seq(0,360, by = 15),
           radial.lim = c(0, max(aq_by_hour$PM2.5.level)),
           add = TRUE)
legend("bottomright", lty = 1, col = c("lightgrey", "blue", "green"), 
       legend = c("PM2.5 level each month", "Mean", "Feb 2016"),
       box.lwd = 0)
###########################################################################
polar.plot(aq_by_hour$Concentration,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "blue", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250),
           # main = paste("Mean PM 2.5 and Concentration", "at each Hour in Beijing", sep = "\n"))
           main = "Monthly Mean Concentration at each Hour")
for (i in year){
  for (j in month){
    polar.plot(aq_monthly_by_hour$Concentration[aq_monthly_by_hour$month == j & 
                                                aq_monthly_by_hour$year == i], 
               start = 90, 
               clockwise = T, 
               rp.type = "polygon",
               line.col = "lightgrey", 
               show.grid.labels = 1,
               labels = aq_monthly_by_hour$hour,
               label.pos = seq(0,360, by = 15),
               radial.lim = c(0, 250),
               lwd = 0.2,
               add = TRUE)
  }
}

polar.plot(aq_by_hour$Concentration,
           start = 90, # move 00 to top
           clockwise = T,
           rp.type = "polygon", # connect 23 to 00
           line.col = "red", 
           show.grid.labels = 1, # (1-4) 1 = move labels to vertical axis on bottom
           labels = aq_by_hour$hour, # outside labels
           label.pos = seq(0,360, by = 15), 
           radial.lim = c(0, 250), # defaults to min and max
           add = TRUE)
polar.plot(aq_monthly_by_hour$Concentration[aq_monthly_by_hour$month == "2" & aq_monthly_by_hour$year == "2016"], 
           start = 90, 
           clockwise = T, 
           rp.type = "polygon",
           line.col = "black", 
           show.grid.labels = 1,
           labels = aq_monthly_by_hour$hour,
           label.pos = seq(0,360, by = 15),
           radial.lim = c(0, max(aq_by_hour$PM2.5.level)),
           add = TRUE)
legend("bottomright", lty = 1, col = c("lightgrey", "red", "black"), 
       legend = c("PM2.5 con. each month", "Mean", "Feb 2016"),
       box.lwd = 0)
```


### Mean PM2.5 Level by Month
```{r}
aq_by_month <- aq_by_month %>%
    mutate(yrmonth =  factor(yrmonth, levels = yrmonth_order)) %>%
   arrange(yrmonth)
plot(aq_by_month$PM2.5.level,
     type = "b",
     axes = FALSE,
     xlab = "Year-Month",
     ylab = "PM2.5",
     main="Mean PM 2.5 Level by Month",
     ylim = c(0,250),
     pch = 16,
     col = "grey")
  axis(1, at= seq(1,73,by=2), 
       tck = FALSE,
       labels = FALSE)
  text(seq(1,73,by=2), par("usr")[3]-0.5, 
       srt = 60, adj= 1, xpd = TRUE,
       labels = paste(levels(aq_by_month$yrmonth)[seq(1,73,by=2)]), 
       cex=0.8)
  axis(2, at= seq(0,(max(aq_by_month$PM2.5.level)+50),by=50),
       tck = 0, las = 1,
       labels = seq(0,(max(aq_by_month$PM2.5.level)+50),by=50))
  par(xpd = FALSE)
  grid()
   points(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2016-2"],
        y = min(aq_by_month$PM2.5.level),
        col = "red",
        pch = 16)
   points(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2011-1"],
         y = aq_by_month$PM2.5.level[aq_by_month$yrmonth=="2011-1"],
         col = "blue",
         pch = 16)
   text(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2015-10"],
       y = min(aq_by_month$PM2.5.level)-20,
       paste("PM2.5 = ", round(min(aq_by_month$PM2.5.level),2)," in Feb 2016", sep = "\n"),
       adj = c(0,1),
       offset = c(0,-2))
   text(x = aq_by_month$yrmonth[aq_by_month$yrmonth =="2011-2"],
       y = aq_by_month$PM2.5.level[aq_by_month$yrmonth=="2011-1"]-20,
       paste("PM2.5 = ", round(aq_by_month$PM2.5.level[aq_by_month$yrmonth=="2011-1"],2), 
             "in Jan 2011", sep = "\n"),
       adj = c(0,1))
```


